---
format:
  revealjs:
    theme: [night, assets/phdcn_style.scss]
    logo: img/phdcn_logo_white.svg
    incremental: false
    self-contained: false
    width: 1200
    height: 800
editor: source
mouse-wheel: true
history: false
---



## [Life-course Processes as Cause and Effect of Social Change in Gun Violence]{.r-fit-text} {background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity="0.3"}


&nbsp;

|                    |                           |
|-------------------:|:--------------------------|
| Charles C. Lanfear | *University of Cambridge* |
| Robert J. Sampson  | *Harvard University*      |



```{r setup}
#| include: false
knitr::opts_chunk$set(dev = "ragg_png",  
                      dev.args = list(bg = 'transparent'),
                      message = FALSE,
                      echo = FALSE,
                      warning = FALSE,
                      fig.showtext = TRUE,
                      fig.retina = 5,
                      fig.align = "center")
library(tidyverse)
library(ggforce)
library(showtext)
library(janitor)
library(flextable)
library(ggtext)
plot_font <- "Open Sans"
font_add_google(name = plot_font)
showtext_auto()
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}
```


## Homicide 1995-2021

```{r homicide-in-chicago, cache = TRUE}
load("./data/hom_rates_race_df.RData")
hom_rates_race_df %>%
  filter(race != "Other") %>%
  ggplot(aes(x = year, y = value, group = race, color = race)) + 
  geom_line(data = ~ . |> filter(type == "Gun Homicide"), linetype = "solid")  +
  geom_line(data = ~ . |> filter(type == "All Homicide"), linetype = "dashed")  +
  geom_richtext(data =  ~ . |> 
                  group_by(race) %>% 
                  filter(year == 1992) |>
                  summarize(value = mean(value), year = 1990.5), 
                aes(label = race), fill = "black") +
  geom_hline(yintercept = 0, color = "white", linewidth = 0.25) +
  annotate("text", x = c(1990.5, 1990.5), y = c(10, 12), label = c("Gun Homicide", "All Homicide"), color = "white", family = plot_font) +
  geom_line(data = tibble(x = c(1993, 1994, 1993, 1994), 
                          y = c(10, 10, 12, 12), 
                          type = factor(c("Gun Homicide", "Gun Homicide", "All Homicide", "All Homicide"), levels = c("Gun Homicide", "All Homicide"))),
            aes(x = x, y = y, linetype = type), color = "white", inherit.aes=FALSE) +
  scale_y_continuous(breaks = seq(0, 30, by = 10)) +
  coord_cartesian(clip = "off", ylim = c(0, 30), xlim = c(1990, 2021)) +
  xlab(NULL) + ylab("Rates per 100,000") +
  theme_minimal(base_size = 24) +
  theme(panel.grid = element_blank(),
        axis.text = element_text(color = "white"),
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        legend.position = "none")
```

Homicide is increasingly committed with guns

## [Handgun violence over the life course]{.r-fit-text}{background-image="img/stephan-cassara-KnAIsBuitGg-unsplash.jpg" background-opacity="0.1"}


```{r handgun-homicide}
load("./data/shr_props_year.RData")
col_vec <- setNames(c("#00BFC4", "#F8766D", "grey50"), c("Adolescence", "Early\nAdulthood", ""))
shr_props_year <- shr_props_year |>
  filter(between(age, 10, 60)) |>
  mutate(stage = case_when(
    between(age, 10, 20) ~ "Adolescence",
    between(age, 21, 45) ~ "Early\nAdulthood",
    TRUE ~ ""
  ))

shr_sums <- shr_props_year |> 
  summarize(pr = sum(pr),
            label = fmt_pct(sum(pr)),
            .by = c(year, stage)) |>
  filter(stage != "") |>
  mutate(x = ifelse(stage == "Adolescence", 10, 35),
         y = 0.04)
shr_medians <- shr_props_year |>
  group_by(year) |>
  filter(cum_pr >= 0.5) |>
  slice_head(n=1) |>
  select(year, age, yend = pr)

shr_props_year |>
ggplot(aes(x = age, y = pr, fill = stage)) + 
  geom_col() +
   geom_text(data = tibble(stage = names(col_vec)[1:2], x = c(8, 38), y = c(0.035, 0.035), year = c(1995,1995)), 
            aes(x=x, y=y, label = stage, color = stage), family = plot_font, size = 9, lineheight = 0.4) +
  geom_text(data = shr_sums, 
            aes(x = x, y = y, label = label, color = stage), size = 6.25, family = plot_font) +
  scale_fill_manual(values = col_vec) +
  geom_segment(data = shr_medians, aes(x = age, xend = age, yend =yend), y = 0, color = "yellow", inherit.aes = FALSE) +
  coord_cartesian(xlim = c(5, 65), ylim = c(0, 0.061), expand = FALSE)  +
  scale_color_manual(values = col_vec) +
  scale_x_continuous(breaks = seq(10, 60, by = 10)) +
  scale_y_continuous(breaks = NULL) +
  facet_wrap(~year, nrow = 1) +
  theme_minimal(base_size = 24) +
  labs(x = "Age of handgun homicide offenders in UCR SHR 1995 and 2021", y = NULL) +
  theme(panel.grid = element_blank(),
        strip.text = element_text(family = plot_font,  color = "white"),
        axis.ticks.x = element_line(color = "white"),
        axis.text = element_text(color = "white"),
        axis.title.x = element_text(color = "white", size = rel(0.8)),
        text = element_text(family = plot_font,  color = "white"),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        plot.margin = margin(0,0,0,0, "pt"),
        legend.position = "none")
```

Homicide offenders are getting older

## Questions {background-image="img/clay-banks-nsAn3nSW5T0-unsplash.jpg" background-opacity="0.3"}

* How does macro-context impact carrying across life course?

. . .

* How does life-course variation in carrying impact rates of violence?


## {.nostretch background-image="img/clay-banks-nsAn3nSW5T0-unsplash.jpg" background-opacity="0.1"}

![](img/phdcn_logo_white.svg){width=700px align=center}

* Cohorts born 1981, 1984, 1987, and 1996
   * Average ages 25 to 40 in 2021

* Key measures:
   * **Onset**: Age first carried a concealed gun
   * **Continuity**: Carried in past year (wave 5 only)
   * **Exposure**: Age first was shot or saw someone else shot

* Survey and attrition weights

# Onset {background-image="img/pedro-lastra-5Bv3EcijAl4-unsplash.jpg" background-opacity="0.3"}

When does **onset** of concealed carry occur?

## [Onset of concealed carry]{.r-fit-text}{background-image="img/pedro-lastra-5Bv3EcijAl4-unsplash.jpg" background-opacity="0.1"}


```{r, eval = FALSE}
load("../../data/derived/survival_data.RData")

draw_curve <- function(x, noaxis=FALSE){
  plot_out <- plot_age_curve(x, "concealed_carry", min_at_risk = 10, weights = TRUE, w5_sample = FALSE, trim_weights = 0.95, intervals = FALSE, font_fam = plot_font) +
  labs(title = str_to_title(x)) +
  geom_vline(xintercept = 21, linetype = "dashed", color = "white", linewidth = 0.5) +
  geom_text(data = ~. |> 
                    distinct(strata) |>
                    mutate(age = case_match(as.character(strata),
                             "Black" ~ 28,
                             "White" ~ 33,
                             "Hispanic" ~ 16,
                             "Male" ~ 26,
                             "Female" ~ 26,
                             "1981" ~ 32,
                             "1987" ~ 29,
                             "1996" ~ 29,
                             "1984" ~ 38
                           ),
                           cdf = case_match(as.character(strata),
                             "Black" ~ 30,
                             "White" ~ 4,
                             "Hispanic" ~ 18.5,
                             "Male" ~ 29,
                             "Female" ~ 8,
                             "1981" ~ 32,
                             "1987" ~ 25,
                             "1996" ~ 3,
                             "1984" ~ 10
                           )),
                  aes(x = age, y = cdf, label = strata, color = strata),
            size = 7,
            family = plot_font) +
  scale_y_continuous(limits = c(0, 60), labels = scales::label_percent(scale = 1)) +
  theme_minimal(base_size = 20) +
  scale_x_continuous(limits = c(5, 45), breaks = seq(10, 40, by = 10)) +
  theme(panel.grid = element_blank(),
        panel.grid.major.y = element_line(color = "grey25"),
        axis.ticks.x = element_line(color = "white"),
        axis.text = element_text(color = "white"),
        text = element_text(family = plot_font,  color = "white"),
        legend.position = "none",
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        plot.margin = margin(0,1,0,1))
  if(noaxis){
    plot_out <- plot_out + theme(axis.text.y = element_blank())
  }
  return(plot_out)
}
(draw_curve("race") + ggtitle("Race / Ethnicity"))+ draw_curve("sex", noaxis = TRUE) + draw_curve("cohort", noaxis = TRUE) &
  plot_annotation(theme = theme(plot.background = element_rect(fill = "transparent",colour = NA), text = element_text(family = plot_font,  color = "white")), caption = "Estimated cumulative incidence from non-parametric Turnbull MLE")
```

* Rapid rise in adolescence
* Steady rise throughout early adulthood

## [Age of first gun use by carry onset]{.r-fit-text}{background-image="img/pedro-lastra-5Bv3EcijAl4-unsplash.jpg" background-opacity="0.1"}


```{r, eval = FALSE}
load("./data/cum_gun_use.RData")
cum_gun_use |>
    mutate(cdf = cdf*100) |>
    ggplot(aes(x = age, y = cdf, color = strata), linewidth = 0.25) +
    geom_line(linewidth = 1) + 
    geom_text(data = ~. |> 
                    distinct(strata) |>
                    mutate(age = case_match(as.character(strata),
                             "Onset before 21" ~ 23,
                             "Onset after 21" ~ 30,
                             "Never carried" ~ 34
                           ),
                           cdf = case_match(as.character(strata),
                             "Onset before 21" ~ 44,
                             "Onset after 21" ~ 22,
                             "Never carried" ~ 7
                           )),
                  aes(x = age, y = cdf, label = strata, color = strata),
            family = plot_font,
            size = 8) +
    labs(x="Age first used gun for 'defense' or shot / shot at someone", y = NULL, title = NULL, color = NULL) +
    scale_x_continuous(limits = c(5, 45), breaks = seq(10, 40, by = 10)) +
    scale_y_continuous(limits = c(0, 50), breaks = seq(0, 40, by = 10),  labels = scales::label_percent(scale = 1)) +
    theme_minimal(base_size = 24) +
    theme(panel.grid = element_blank(),
        panel.grid.major.y = element_line(color = "grey25"),
        axis.ticks.x = element_line(color = "white"),
        axis.title.x = element_text(size = rel(0.8)),
        axis.text = element_text(color = "white"),
        text = element_text(family = plot_font,  color = "white"),
        legend.position = "none",
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        plot.margin = margin(0,1,0,1))
```

* 42% of young carriers used gun by 21
    * Those who didn't never used a gun later in life
* Lifetime gun use nearly equal by age 40


# Continuity {background-image="img/benjamin-suter-mpLex62zVKQ-unsplash.jpg" background-opacity="0.3"}

Is there **continuity** in carry over the life course?


## [Past-year carry in 2021]{.r-fit-text}{background-image="img/benjamin-suter-mpLex62zVKQ-unsplash.jpg" background-opacity="0.1"}


```{r, eval = FALSE}
load("../../data/derived/continuity_plot_data.RData")
continuity_plot_data |> 
  mutate(Initiation = factor(ifelse(pre_21, "Onset\nbefore\n21", "After\n21"), levels = c("Onset\nbefore\n21", "After\n21")),
         concealed_carry_status = 
           fct_recode(concealed_carry_status, 
                      `No carry\nin past year` = "no carry",
                      `Carry with\na permit` = "carry with permit",
                      `Carry without\na permit` = "carry without permit"))  |>
  count(Initiation, concealed_carry_status) |>
  mutate(pr = n / sum(n), .by = Initiation) |>
  ggplot(aes(x = concealed_carry_status, group = Initiation, fill = Initiation, y = pr)) +
  geom_col(position = "dodge") +
   geom_text(data = ~ . |> distinct(Initiation) |>
               mutate(pr = 0.02,
                      x = ifelse(Initiation == "Onset\nbefore\n21", .77, 1.23)),
            aes(x = x, label = Initiation, y = pr), color = "black", vjust = 0, hjust = 0.5, size = 9, lineheight = .8,
            family = plot_font) +
  theme_minimal(base_size = 24) +
  coord_cartesian(ylim = c(0, 0.7)) +
  scale_y_continuous(limits = c(0, .70), labels = scales::label_percent(scale = 100)) +
  # annotate("text", x = 0, y = 0.7, label = "Age", size = 9,  hjust = 0, color = "white") +
  labs(x = NULL, y = NULL, caption = "Excluding those with onset within past year") +
  theme(panel.grid = element_blank(),
        text = element_text(family = plot_font,  color = "white"),
        axis.ticks.length = unit(0, "pt"),
        plot.caption = element_text(size = rel(0.5)),
        axis.text = element_text(color = "white"),
        axis.text.x = element_text(vjust = 0.5, margin = margin(0,0,0,0), size = rel(1.2)),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        panel.grid.major.y = element_line(color = "grey25"),
        legend.position = "none",
        plot.margin = margin(0,0,0,0))
```

* 60% of young carriers not carrying today
* Young carriers no more likely to carry without permits




## New Questions {background-image="img/joel-mott-s-rsM-AktbA-unsplash.jpg" background-opacity="0.1"}




## Feedback and Questions {.smaller background-image="img/joel-mott-s-rsM-AktbA-unsplash.jpg" background-opacity="0.1"}

Contact:

| Charles C. Lanfear
| Institute of Criminology
| University of Cambridge
| [cl948\@cam.ac.uk](mailto:cl948@cam.ac.uk)

&zwj;

For more about the PHDCN+:

| PHDCN\@fas.harvard.edu
| <https://sites.harvard.edu/phdcn/>
| [https://doi.org/10.1007/s40865-022-00203-0](https://sites.harvard.edu/phdcn/)

# Appendix {background-image="img/max-bender-yb7Yg3Rv7WA-unsplash.jpg" background-opacity="0.3"} 
